---
title: "Bioinformatics Project - Bulk RNA-seq"
author: "Bidiscombe, Carnevale, Cattoni"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# install recount3 
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("recount3", quietly = TRUE)) {
  BiocManager::install("recount3")
}

library(recount3)
library(SummarizedExperiment)

# 1) List of available projects in recount3
projects <- available_projects()

# 2) we want TCGA (The Cancer Genome Atlas)
tcga_projects <- subset(projects, file_source == "tcga")

# 3) select LUAD (Lung Adenocarcinoma)
projinfo_luad <- subset(tcga_projects, project == "LUAD")
projinfo_luad


# 4) Download the SummarisedExperiment object with the counts (genes x samples)
rse_gene <- create_rse(projinfo_luad)

# 5) Extract sample count matrix and metadata
counts  <- assay(rse_gene, "raw_counts")   # raw count matrix
coldata <- colData(rse_gene)           # metadata (info on samples)

dim(counts)              # n genes x n samples
colnames(counts)[1:5]    

# 6) 2 tipy of smaples: Primary Tumor vs Solid Tissue Normal
sample_type <- coldata$tcga.gdc_cases.samples.sample_type
table(sample_type)

# 7) filter for only tumor and not tumor
keep <- sample_type %in% c("Primary Tumor", "Solid Tissue Normal")
rse_gene_sub <- rse_gene[, keep]

counts_sub  <- assay(rse_gene_sub, "raw_counts")           
coldata_sub <- as.data.frame(colData(rse_gene_sub))   

dim(counts_sub)
table(coldata_sub$tcga.gdc_cases.samples.sample_type)

# 8) Variable 'condition' = normal / tumor
condition <- ifelse(
  coldata_sub$tcga.gdc_cases.samples.sample_type == "Primary Tumor",
  "tumor",
  "normal"
)

coldata_sub$condition <- factor(condition, levels = c("normal", "tumor"))

# Final check
table(coldata_sub$condition)
```
PART 2: PPI Network Analysis

This section implements the Protein-Protein Interaction analysis as required by the project guidelines, replicating the logic of the provided Python notebook.

2.1 Load and Clean PPI Data (HIPPIE)

We use the HIPPIE database (v2.3). We filter for high-confidence interactions ($\ge 0.65$) and map identifiers to Gene Symbols.
```{r}
library(tidyverse)

# Loading of the downloaded HIPPI file
possible_paths <- c("data/HIPPIE-current.mitab.txt", "HIPPIE-current.mitab.txt")
found_path <- NULL

for (path in possible_paths) {
  if (file.exists(path)) {
    found_path <- path
    break
  }
}

if (is.null(found_path)) {
  stop("File not found")
} else {
  message(paste("File found in:", found_path))
}

# Loading data
hippie_cols <- c(
  "ID_Interactor_A", "ID_Interactor_B", "Alt_IDs_Interactor_A", "Alt_IDs_Interactor_B",
  "Aliases_Interactor_A", "Aliases_Interactor_B", "Interaction_Detection_Methods",
  "Publication_1st_Author", "Publication_Identifiers", "Taxid_Interactor_A", "Taxid_Interactor_B",
  "Interaction_Types", "Source_Databases", "Interaction_Identifiers", "Confidence_Value",
  "Presence_In_Other_Species", "Gene_Name_Interactor_A", "Gene_Name_Interactor_B"
)

hippie <- read_tsv(found_path, col_names = hippie_cols, comment = "#", show_col_types = FALSE)

# only confidence >= 0.65, other removed
edges_clean <- hippie %>%
  filter(Confidence_Value >= 0.65) %>%
  mutate(
    ID_Interactor_A = str_remove(ID_Interactor_A, "entrez gene:"),
    ID_Interactor_B = str_remove(ID_Interactor_B, "entrez gene:")
  ) %>%
  select(ID_Interactor_A, ID_Interactor_B, Confidence_Value)

cat("Interactions loaded (Score >= 0.65):", format(nrow(edges_clean), big.mark=","), "\n")
```

## 2.2 Gene Annotation (NCBI Mapping)

To ensure data quality, we validated all HIPPIE interactions 
against the NCBI human gene database. This filtering step:

1. Removes interactions involving deprecated or invalid gene identifiers
2. Ensures bidirectional mapping between Entrez IDs and Gene Symbols
3. Guarantees biological relevance of all protein interactions

Result: 206,000 validated interactions (vs 207,000 original HIPPIE interactions),
representing a 99.5% retention rate with improved data integrity.

```{r part2_ncbi}
library(tidyverse)

# FILE LOCALI NCBI
ncbi_filename <- "Homo_sapiens.gene_info.gz"
possible_ncbi_paths <- c(paste0("data/", ncbi_filename), ncbi_filename)
found_ncbi_path <- NULL

for (path in possible_ncbi_paths) {
  if (file.exists(path)) {
    found_ncbi_path <- path
    break
  }
}

# Only columns that wre need
genes_ncbi <- read_tsv(found_ncbi_path, show_col_types = FALSE) %>%
  select(GeneID, Symbol, description) %>%
  mutate(GeneID = as.character(GeneID))

# keep interactions in the network where both genes are recognised
edges_final <- edges_clean %>%
  filter(ID_Interactor_A %in% genes_ncbi$GeneID & ID_Interactor_B %in% genes_ncbi$GeneID)

cat("Final Valid Interactions:", format(nrow(edges_final), big.mark=","), "\n")
```

## 2.3 Network Construction and Analysis (igraph)

We construct the PPI graph and analyse the Giant Component (the largest connected part of the network).

```{r part2_network}
library(igraph)

# Graph construction (Network)
ppi_net <- graph_from_data_frame(d = edges_final, directed = FALSE)
# 'edges_final' contains pairs of interacting genes

# add symbols to nodes in the network.
V(ppi_net)$symbol <- genes_ncbi$Symbol[match(V(ppi_net)$name, genes_ncbi$GeneID)]

# remove self-links and duplicates
ppi_net <- simplify(ppi_net, remove.multiple = TRUE, remove.loops = TRUE)

# extraction of Giant Component
comps <- components(ppi_net)
giant_id <- which.max(comps$csize)
giant_component <- induced_subgraph(ppi_net, which(comps$membership == giant_id))

cat("Nodes:", format(vcount(giant_component), big.mark=","), "\n")

```
## 2.4 Topology Analysis: Hub Identification

We calculate the degree distribution to confirm the scale-free nature of the network and identify the most connected proteins (Hubs).

```{r part2_hubs, fig.width=8, fig.height=5}
library(ggplot2)
library(knitr)
library(dplyr)

# 1. Calculation of degree (n connections x node)
node_degrees <- degree(giant_component)
deg_df <- data.frame(GeneID = names(node_degrees), Degree = node_degrees)

# 2. Grade sitribution chart (log-log Scale)
ggplot(deg_df, aes(x = Degree)) +
  geom_histogram(bins = 50, fill = "darkblue", color = "white", alpha=0.8) +
  scale_y_log10() + 
  scale_x_log10() +
  labs(title = "degree distribution (log-log Scale)", 
       subtitle = "Network Topology Analysis",
       x = "Degree (k)", y = "Frequency P(k)") +
  theme_minimal()

# 3. identification ofi Top 15 Hubs
top_hubs <- deg_df %>%
  arrange(desc(Degree)) %>%
  head(15) %>%
  left_join(genes_ncbi, by = "GeneID") %>%
  select(Symbol, Degree, description)

kable(top_hubs, caption = "Top 15 Hub Genes identified in the Network")
```
We calculate the degree distribution to confirm the scale-free nature of the network and identify the most connected proteins (Hubs).

### Degree Distribution Analysis

The degree distribution plot see the relationship between the number of connections (degree on the X axes) and their frequency (Y axes) on a log-log scale. This visualization reveals the **scale-free network topology** characteristic of biological protein-protein interaction networks:

- **Power-law distribution**: The decreasing trend on the log-log plot indicates that the network follows a power-law distribution, typical of scale-free networks.
- **Hub nodes**: A small number of highly connected proteins (hubs) dominate the network, while most proteins have relatively few interactions.
- **Biological significance**: Hub proteins are often essential regulators or central players in cellular processes, making them important targets for disease intervention.

The presence of highly connected nodes like ESR1 and CUL3 suggests these proteins are central coordinators of biological pathways.

**Key observations:**
- **TP53** and **MYC** are known tumor suppressors and proto-oncogenes, central to lung cancer biology.
- **EGFR** is a major target for lung cancer therapeutics.
- Cullin proteins (CUL3, CUL1) are ubiquitination machinery components essential for protein degradation pathways.
- The extreme connectivity of these hubs (e.g., ESR1 with 1,963 interactions) reflects their role as integration points in cellular signaling.
-----------------------------------------------------------------------------------



## 2.5 Network Propagation (Random Walk with Restart)

We use the Random Walk with Restart (RWR) algorithm to propagate the signal from the Differentially Expressed (DE) genes to the rest of the network. This allows us to identify **hidden genes** that are not mutated themselves but interact closely with disease genes.

```{r part2_propagation}
library(igraph)

# ----- Simulation part for when the data from the prev section were not avaible
#set.seed(123)
#disease_gene_ids <- sample(V(giant_component)$name, 50)
# --- end SIMULAZIONE

# 1. load the file CSV 
de_results <- read.csv("data/results_de_genes_LUAD.csv")
#
# 2. Filter significant genes (padj < 0.05)
# disease_gene_ids <- as.character(de_results$entrez_id[de_results$padj < 0.05])
de_results_filtered <- de_results %>%
  filter(!is.na(entrez_id)) %>%
  filter(FDR < 0.05)#
disease_gene_ids <- as.character(de_results_filtered$entrez_id)

# 3. Ensure that these genes are present in the network
# disease_gene_ids <- intersect(disease_gene_ids, V(giant_component)$name)
disease_gene_ids <- intersect(disease_gene_ids, V(giant_component)$name)
# ===== CHECK: se nessun gene è nel network, ferma e stampa debug =====
if (length(disease_gene_ids) == 0) {
  stop("ERRORE: Nessun gene DE trovato nella rete PPI!")
}

# ALGORITHM EXECUTION (Equivalent to random_walker in the Python notebook)
personalized_vector <- rep(0, length(V(giant_component)))
names(personalized_vector) <- V(giant_component)$name
personalized_vector[disease_gene_ids] <- 1
personalized_vector <- personalized_vector / sum(personalized_vector)

# 85% prob restart
rwr_scores <- page_rank(
  giant_component, 
  directed = FALSE, 
  damping = 0.85, 
  # personalized = as.numeric(V(giant_component)$name %in% disease_gene_ids)
  personalized = personalized_vector
)$vector

# Creating Results Table
results_df <- data.frame(
  GeneID = names(rwr_scores),
  Score = rwr_scores,
  Symbol = V(giant_component)$symbol
) %>%
  arrange(desc(Score))

# We identify NEW candidates
# (High-scoring genes that were NOT on the initial list of diseased genes)
new_candidates <- results_df %>%
  filter(!GeneID %in% disease_gene_ids) %>%
  head(100) # Prendiamo i top 100

cat("Top 10 New Candidates identified by Network Propagation:\n")
kable(head(new_candidates[, c("Symbol", "Score")], 10))
```
### 2.5b Enriched Disease Module Visualization
```{r}
library(igraph)
library(dplyr)

# Get top 13 genes 
top_15_genes <- results_df %>% head(13) %>% pull(GeneID)
enriched_subgraph <- induced_subgraph(giant_component, top_15_genes)

V(enriched_subgraph)$type <- ifelse(V(enriched_subgraph)$name %in% disease_gene_ids, 
                                     "Seed", "Predicted")
V(enriched_subgraph)$label <- V(enriched_subgraph)$symbol

cat("Enriched Module - Top 13 Hub Genes:\n")
cat("Nodes:", vcount(enriched_subgraph), "\n")
cat("Edges:", ecount(enriched_subgraph), "\n\n")

# Layout
set.seed(123)
layout <- layout_with_fr(enriched_subgraph, 
                         niter = 1500,
                         start.temp = 150)
layout <- layout * 30

colors <- ifelse(V(enriched_subgraph)$type == "Seed", "red", "lightblue")
sizes <- (degree(enriched_subgraph) + 1) * 3.5

plot(enriched_subgraph,
     vertex.size = sizes,
     vertex.color = colors,
     vertex.label.cex = 1.2,
     vertex.label.dist = 1,
     vertex.label.family = "sans",
     edge.width = 2,
     edge.color = rgb(0, 0, 0, 0.2),
     main = "Enriched Disease Module - Top 15 Hub Genes",
     sub = "Red: DE genes (seed) | Blue: Network candidates",
     layout = layout)

legend("topright", 
       legend = c("Seed (DE gene)", "Predicted candidate"),
       col = c("red", "lightblue"),
       pch = 19,
       cex = 1.2,
       pt.cex = 2)
```
The enriched disease module reveals the top 15 most highly-scored genes from the RWR analysis. Red nodes represent seed genes (differentially expressed), while blue nodes represent novel candidates predicted by network propagation. 

**Key findings:**
- **PCNA** emerges as a central hub, with downregulation in tumor tissue
- **TP53** and **MYC** are predicted candidates highly connected to multiple pathways
- The clustering suggests dysregulation of DNA replication and cell cycle control
- Hub proteins like EGFR and HSP family members represent potential therapeutic targets



## 2.6 Functional Enrichment Analysis

We perform pathway enrichment analysis on the top 100 predicted genes using `gprofiler2` to understand their biological functions (GO, KEGG, Reactome).

```{r part2_enrichment, fig.width=10, fig.height=8}
library(gprofiler2)

# Enrichment
gostres <- gost(query = new_candidates$Symbol, 
                organism = "hsapiens", 
                ordered_query = TRUE, 
                significant = TRUE,
                sources = c("GO:BP", "KEGG", "REAC"))

if (!is.null(gostres$result)) {
  # 1. Manhattan Plot 
  p <- gostplot(gostres, capped = FALSE, interactive = FALSE)
  print(p)
  
  # 2. Top Pathway
  top_pathways <- gostres$result %>%
    select(source, term_name, p_value, intersection_size) %>%
    arrange(p_value) %>%
    head(15)
    
  kable(top_pathways, caption = "Top Enriched Pathways for Predicted Genes")
} else {
  cat("No significant enrichment found (likely due to random simulated data).\n")
}
```
We perform pathway enrichment analysis on the top 100 predicted genes 
identified by network propagation using g:Profiler2. This analysis integrates 
three major pathway databases: Gene Ontology Biological Processes (GO:BP), 
KEGG pathways, and Reactome.

### Manhattan Plot Interpretation

The Manhattan plot visualizes the statistical significance of pathway enrichments 
across different sources:

- **Orange points (GO:BP)**: Gene Ontology biological processes dominate the results, 
  reflecting the detailed functional annotations available in this database.
- **Pink points (KEGG)**: Metabolic and signaling pathways, showing intermediate 
  significance levels.
- **Blue points (REAC)**: Reactome pathways with the highest statistical significance 
  (p-value < 0.001), indicating robust biological relevance.

### Key Enriched Pathways

The top 15 enriched pathways reveal strong biological relevance:

**Cell Death and Apoptosis (p-value = 0):**
- Programmed cell death (51 genes)
- Apoptotic process (49 genes)
- Cell death (51 genes)

These pathways are highly dysregulated in lung cancer. LUAD cells evade apoptosis 
through alterations in TP53, EGFR, and other predicted hub genes.

**Metabolic Regulation (p-value = 0):**
- Regulation of protein metabolic process (53 genes)
- Regulation of primary metabolic process (77 genes)
- Positive regulation of macromolecule metabolic process (66 genes)

PCNA and predicted hub proteins regulate DNA replication machinery and metabolic 
reprogramming in cancer cells.

**Signal Transduction and Stress Response:**
- Intracellular signal transduction (59 genes)
- Cellular response to stress (47 genes)

These pathways highlight the dysregulation of growth signaling and cellular stress 
responses in LUAD.

**Reactome Disease Pathway (p-value = 0):**
- 64 genes directly annotated as disease-associated, providing direct validation 
  that our network propagation identified clinically relevant genes.

### Conclusion

The comprehensive enrichment analysis strongly validates the biological relevance 
of the network-predicted candidates. The convergence of multiple pathway databases 
on cancer-relevant processes (apoptosis, metabolic regulation, signal transduction) 
confirms that random walk with restart successfully identified genes involved in 
lung adenocarcinoma pathogenesis.



